01 Calculating Travel Time for TTS16 OD

### 2022-10-22 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
additions to the TTS2016R package and minor edits to the manuscript (more work to be done).

This manuscript presents the open data product [{TTS2016R}](https://github.com/soukhova/TTS2016R). Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open]. The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data. This package includes person-to-jobs origin-destinations, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018]. In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r]. The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required. In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process. By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region. It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future. -->

# Home-to-work commute data

{TTS2016R} includes counts of fully-employed population by place of residence (origin) and counts of full-time jobs by place of work (destination) aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries). TAZ typically are defined based on land-use and population demographics in order to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001]. As such, each TAZ is uniquely identified using the GTA06 Zoning System which can be used to join to the origin-destination table (i.e., trips taken).

The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) in this package are organized in the form of an origin-destination table which is indicative of home-to-work commute patterns (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions). These data were retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].

Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}. These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible applications. For simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid. The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode. Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in Figure \ref{fig:TTS-16-survey-area}.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# grouping the planning boundaries/municipalities so they make up the 20 regions in the TTS 2016. Note: st_buffer is used as there are small existing gaps between some boundaries. st_buffer of 10 m is enough to widen all boundaries and complete the st_union without issue.
group_ggh_pd_poly <- TTS2016R::ggh_pd %>% st_buffer(10) %>% group_by(REGION) %>% 
  summarize(REGION_name = first(REGION_name),
            geometry = st_union((geometry)))

>>>>

This manuscript presents the open data product [{TTS2016R}](https://github.com/soukhova/TTS2016R). Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open]. The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data. This package includes full-time home-based work-to-job origin destinations counts and mode-specific trip numbers, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018]. In addition, the package includes TAZ centroid-to-centroid travel times by car, transit, cyling, and walking mode computed using package {r5r} [@Pereira2021r5r].

The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications. Though data from the TTS is freely available to the public (URL <!-- -->), the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process. By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region and amends calculated travel times to the data set. This package also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.

# Home-to-work commute data
<!-- NOTE: When extracting trip vs. person data in the TTS database. For 'by person' extraction: Trips made by person include all trips. For 'by trip' extraction: trips made do not include walking trips to non-school or non-work purposes, so the total number of trips extracted through this method is NOT equal to the total number of trips per person (the total number of trips per person is always higher than the total number of trips). You can't filter the 'by person' extraction by trip purpose. You can filter the 'by trip' extraction by trip purpose, so we will be using the 'by trip' extraction - filtered by 'home-based work' - to get the primary mode trip numbers. -->
{TTS2016R} includes counts of full-time employed population by place of residence (origin), counts of full-time usual place of work (destination), number of trips to work by mode, average estimated travel time of trip, and the number of vehicles by mode. This data is aggregated and available at the level of TAZ: TAZ are a spatial unit of analysis typically used to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001]. They are thus defined by transportation planners for a region based on intra-similarity and inter-dissimilarity between land-use and population demographics. Within the GGH boundaries, `r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` TAZ are specified. The TAZ range between ≥`r 0.019` km^2 in spatial area to a maximum of `r max(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 (median: `r median(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 and 3rd quantile: `r quantile(TTS2016R::ggh_taz$AREA, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")` km^2).

The number of full-time working population (will be referred to as 'workers' throughout) and full-time places of employment (will be referred to as 'jobs' throughout) is also quite variable between TAZ. In terms of workers, TAZ contain between `r min(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` workers (median: `r median(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$workers, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`). In terms of jobs, TAZ contain between `r min(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` jobs (median: `r median(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$jobs, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`).

Each TAZ is uniquely identified using the GTA06 Zoning System which can be used to match the worker residential location to the job location in the origin-destination table. The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) in this package are organized in the form of an origin-destination table which is indicative of home-to-work commute patterns (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions). This data was retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the interaction of full-time employed people to their places of employment as shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].

Also included in {TTS2016R} are the estimated travel times from origin to destination by various modes; travel times are calculated using the package {r5r} and the notebook for this calculation is availalbe in the documentation of {TTS2016R}. These travel times are useful addition to {TTS2016R} since they are not included in the TTS Data Retrieval System but they are vitally important to estimate the cost of travel and associated impedance functions, among other possible applications. The included travel times are by car, walking, cyling, and public transit modes and are calculated from each origin TAZ centroid to each destination TAZ centroid where a home-based work trip is made (`r TTS2016R::od_ft_tt$Origin %>% length() %>% prettyNum(big.mark = ",") ` OD pairs). The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all pairs. Additionally, for car and transit modes only travel times less than or equal to 180 mins are calculated. For cycling and walking modes only travel times less than or equal to 120 mins are calculated.

These threshold represents 90% or greater of trip's travel times, per mode, which are summarized in Figure \ref{fig:TTS-16-survey-area}.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# grouping the planning boundaries/municipalities so they make up the 20 regions in the TTS 2016. Note: st_buffer is used as there are small existing gaps between some boundaries. st_buffer of 10 m is enough to widen all boundaries and complete the st_union without issue.
group_ggh_pd_poly <- TTS2016R::ggh_pd %>% st_buffer(10) %>% group_by(REGION) %>% 
  summarize(REGION_name = first(REGION_name),
            geometry = st_union((geometry)))

### 2022-10-22 data-raw/00-TTS16-and-Bound-Data-Preparation-OD.Rmd
additions to the TTS2016R package and minor edits to the manuscript (more work to be done).

Load packages:
```{r load-packages}
library(readxl)
library(sf)
library(tidyverse)
library(units)
```

@@

Read the boundaries of the traffic analysis zones:
```{r}
ggh_taz <- st_read("TTS16-data-inputs/OD-by-FT-Employment/tts06_83_region.shp")
```

@@

Read the boundaries of the traffic analysis zones:
```{r}
ggh_pd <- st_read("TTS16-data-inputs/Boundaries/tts06_pd_83.shp")
```

@@

## Origin destination matrices for work and shop

Read trip tables. The query was for "trips" by zone of residence, zone of work, and trip purpose (Work):
```{r}
od_ft <- read_delim(file = "TTS16-data-inputs/OD-by-FT-Employment/tts-2016-OD-FT-Employment.txt", 
                      delim = "\t", 
                      col_names = FALSE)
```

@@

@@

The od table can be used to calculate the workers and jobs by traffic analysis zone:
```{r}
workers_ft <- od_ft %>%
  group_by(Origin) %>%
  summarize(workers = sum(Persons),
            .groups = "drop") %>%
  transmute(GTA06 = Origin,
            workers)

jobs_ft <- od_ft %>%
  group_by(Destination) %>%
  summarize(jobs = sum(Persons),
            .groups = "drop") %>%
  transmute(GTA06 = Destination,
            jobs)

Join workers and jobs to simple features object:
```{r}
ggh_taz <- ggh_taz %>%
  # Join workers
  left_join(workers_ft,
            by = "GTA06") %>%
  # Join jobs
  left_join(jobs_ft,
            by = "GTA06") %>%
  # Replace NAs with zeros
  mutate(workers = replace_na(workers, 0),
         jobs = replace_na(jobs, 0))
```

@@

Plot job distribution in the ggh:
```{r}
ggplot() +
  geom_sf(data = ggh_taz,
          aes(fill = jobs),
          color = NA) +
  scale_fill_fermenter(direction = 1)
```

@@

## saving inputs for later processing
```{r}
save("ggh_taz", file = "TTS16-data-inputs/OD-by-FT-Employment/ggh_taz.Rdata")
save("ggh_pd", file = "TTS16-data-inputs/Boundaries/ggh_pd.Rdata")
save("od_ft", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft.Rdata")
```

>>>>

Load packages:
```{r load-packages}
library(readxl)
library(sf)
library(tidyverse)
library(units)
library(tmap)
library(ggplot2)
```

@@

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_taz <- st_read("TTS16-data-inputs/OD-by-FT-Employment/tts06_83_region.shp")
```

@@

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_pd <- st_read("TTS16-data-inputs/Boundaries/tts06_pd_83.shp")
```

@@

add 'REGION' from the planning distract object to the `ggh_taz`. 
```{r}
#buffering to clean up the edges of these objects -> there wre issues using st_join. 
ggh_taz <- st_buffer(ggh_taz, dist=0)
ggh_pd <- st_buffer(ggh_pd, dist=0)

ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
```

## CMA and CA boundaries
If there's interest in tying the TAZ to the Canadian Census, the boundaries and IDs for census metropolitan areas (CMA) and census ares (CA) are provided in this package as well. These areas are determined by Statistics Canada and

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_cma <- st_read("TTS16-data-inputs/Boundaries/lcma000b16a_e.shp")
```

Project object and retain specific columns:
```{r}
ggh_cma <- ggh_cma %>%
  st_transform(crs = 32617) %>%
  transmute(CMAUID,
            CMANAME,
            CMATYPE)
```

add 'CMAUID' from the cma object to the `ggh_taz`. 
```{r}
#doing a st_join doesn't provide a clean solution, so I find the centroids of each TAZ, st_join them to the CMA/CA, and then join the resulting CMAUID property to the ggh_taz object
test_pts <-st_centroid(ggh_taz)
test<-st_join(test_pts, ggh_cma %>% transmute(CMAUID), left = TRUE, join=st_within)
ggh_taz <-merge(ggh_taz, test %>% transmute(GTA06, CMAUID)%>% st_drop_geometry(), by=c("GTA06"))
```

## Origin destination matrices for place of residence and typical place work (by person)
Read exported TTS 2016 tables. The query was a cross-tabulation of "persons" by zone of residence and place of employment. The data was filtered to only include full-time not-remote workers and was retrieved from the TTS 2016:
```{r}
od_ft <- read_delim(file = "TTS16-data-inputs/OD-by-FT-Employment/tts-2016-OD-FT-Employment.txt", 
                      delim = "\t", 
                      col_names = FALSE)
```

@@

## Trips taken (by mode) by worker from place of residence to job at place of work
Read exported TTs 2016 tables. The query was a cross-tablulation of "trips" by zone of residence and place of employment. The data was filtered to only include full-time not-remote working trip purposes by primary mode and was retrieved from the TTS 2016:
```{r}
od_ft_trips <- read_delim(file = "TTS16-data-inputs/OD-by-FT-Employment/tts-2016-OD-FT-Employment-by_W_R_trip-byprimmode.txt", 
                      delim = "\t", 
                      col_names = FALSE)
```

Find the position in the table where the information on work trips begins:
```{r}
id_transit <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Transit excluding GO rail)")
id_cycle <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Cycle)")
id_car_driver <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Auto driver)")
id_GO <- which(od_ft_trips$X1 == "TABLE    : mode_prime (GO rail only)")
id_GO_transit <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Joint GO rail and local transit)")
id_motorcycle <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Motorcycle)")
id_other <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Other)")
id_car_pass <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Auto passenger)")
id_schoolbus <- which(od_ft_trips$X1 == "TABLE    : mode_prime (School bus)")
id_taxi <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Taxi passenger)")
id_rideshare <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Paid rideshare)")
id_walk <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Walk)")
id_end <- od_ft_trips$X1 %>% length()
```

Slice the table to obtain the work trips by primary mode (there are 12 possible modes, so 12 new tables) (remove heading information and other modes):
```{r}
od_ft_trips_transit <- od_ft_trips %>% 
  slice((id_transit+3):id_cycle-1)

od_ft_trips_cycle <- od_ft_trips %>% 
  slice((id_cycle+3):id_car_driver-1)

od_ft_trips_car_driver <- od_ft_trips %>% 
  slice((id_car_driver+3):id_GO-1)

od_ft_trips_GO <- od_ft_trips %>% 
  slice((id_GO+3):id_GO_transit-1)

od_ft_trips_GO_transit <- od_ft_trips %>% 
  slice((id_GO_transit+3):id_motorcycle-1)

od_ft_trips_motorcycle <- od_ft_trips %>% 
  slice((id_motorcycle+3):id_other-1)

od_ft_trips_other <- od_ft_trips %>% 
  slice((id_other+3):id_car_pass-1)

od_ft_trips_car_pass <- od_ft_trips %>% 
  slice((id_car_pass+3):id_schoolbus-1)

od_ft_trips_schoolbus <- od_ft_trips %>% 
  slice((id_schoolbus+3):id_taxi-1)

od_ft_trips_taxi<- od_ft_trips %>% 
  slice((id_taxi+3):id_rideshare-1)

od_ft_trips_rideshare <- od_ft_trips %>% 
  slice((id_rideshare+3):id_walk-1)

od_ft_trips_walk <- od_ft_trips %>% 
  slice((id_walk+2):id_end)
```

Let's consolidate the 12 tables into 1 table, based on origin - destination and clean up the table. If there's no trip using that mode, a 0 is added. The following are the codes:

- T_LT - Local transit excluding GO rail
- T_LTGO - Local transit and GO
- T_GO - GO rail excluding local transit
```{r}
od_ft_trips_transit <- od_ft_trips_transit %>%
  separate(X1, into = c("Zones", "T_LT"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         T_LT = as.numeric(T_LT))

od_ft_trips_GO_transit <- od_ft_trips_GO_transit %>%
  separate(X1, into = c("Zones", "T_LTGO"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         T_LTGO = as.numeric(T_LTGO))

od_ft_trips_GO <- od_ft_trips_GO %>%
  separate(X1, into = c("Zones", "T_GO"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         T_GO = as.numeric(T_GO))
```

- BC - Cycle
- W - Walk
```{r}
od_ft_trips_cycle <- od_ft_trips_cycle %>%
  separate(X1, into = c("Zones", "BC"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         BC = as.numeric(BC))

od_ft_trips_walk <- od_ft_trips_walk %>%
  separate(X1, into = c("Zones", "W"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         W = as.numeric(W))
```

- C_DPC - Driver private car 
- C_PPC - Passenger private car
- C_TA - Taxi
- C_RS - Rideshare
- C_SB - School bus
```{r}
od_ft_trips_car_driver <- od_ft_trips_car_driver %>%
  separate(X1, into = c("Zones", "C_DPC"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         C_DPC = as.numeric(C_DPC))

od_ft_trips_car_pass <- od_ft_trips_car_pass %>%
  separate(X1, into = c("Zones", "C_PPC"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         C_PPC = as.numeric(C_PPC))

od_ft_trips_taxi <- od_ft_trips_taxi %>%
  separate(X1, into = c("Zones", "C_TA"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         C_TA = as.numeric(C_TA))

od_ft_trips_rideshare <- od_ft_trips_rideshare %>%
  separate(X1, into = c("Zones", "C_RS"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         C_RS = as.numeric(C_RS))

od_ft_trips_schoolbus <- od_ft_trips_schoolbus %>%
  separate(X1, into = c("Zones", "C_SB"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         C_SB = as.numeric(C_SB))
```

- M - Motorcycle
- O - Other
```{r}
od_ft_trips_motorcycle<- od_ft_trips_motorcycle %>%
  separate(X1, into = c("Zones", "M"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         M = as.numeric(M))

od_ft_trips_other <- od_ft_trips_other %>%
  separate(X1, into = c("Zones", "O"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         O = as.numeric(O))
```

And now, consolidate all 12 tables into 1 table! with 12 columns for mode trips numbers + the origin and destination columns (those are GTA06 TAZ IDs):
```{r}
od_ft_trips <- full_join(od_ft_trips_transit, od_ft_trips_GO_transit, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_GO, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_cycle, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_walk, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_car_driver, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_car_pass, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_taxi, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_rideshare, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_schoolbus, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_motorcycle, by=c("Origin", "Destination")) %>%
  full_join(od_ft_trips_other, by=c("Origin", "Destination")) 
```

Replace all NAs with 0s:
```{r}
od_ft_trips[is.na(od_ft_trips)] <- 0
```

Combine the `od_ft` person data frame with the `od_ft_trips` data frame. Notice that there are more OD pairs in the `od_ft` data frame (`r od_ft$Origin %>% length()`) than `od_ft_trips` (`r od_ft_trips$Origin %>% length()`) because some OD pairs which have workers and/or are a location for jobs do not have any trips taken that survey day. The TTS 2016 is a travel survey after all. 
```{r}
od_ft_p_t <- full_join(od_ft, od_ft_trips, by=c("Origin", "Destination")) 
```

Check to see if the total number of trips adds up to the total available from TTS 2016 (to ensure we separated and merged all the tables correctly)
```{r}
od_ft_p_t <- od_ft_p_t %>% rowwise() %>%
  mutate(sum_trips = sum(T_LT,T_LTGO,T_GO,BC,W,C_DPC,C_PPC,C_TA,C_RS,C_SB,M,O, na.rm=T))

#the number of total trips. This value is slightly higher than the value for total trips (not disagg. by mode, which is equal to 3,282,539 - I checked in excel) as available from the TTS 2016. However, I checked the summed up total trip, disaggregated by mode, in excel, and it yields this number of 3,282,611 as well. So there must just be some rounding up that occurred in the TTS 2016 when disaggregating by mode... regardless the numerical difference is insignificant. 
od_ft_p_t$sum_trips %>% sum()
```

Replace all NAs with 0s, again:
```{r}
od_ft_p_t[is.na(od_ft_p_t)] <- 0
```

@@

The od table can be used to calculate the count of workers, count of jobs, and count of % transit, bike, and walk slit, by traffic analysis zone:
```{r}
workers_ft <- od_ft_p_t %>%
  group_by(Origin) %>%
  summarize(workers = sum(Persons),
            .groups = "drop") %>%
  transmute(GTA06 = Origin,
            workers)

jobs_ft <- od_ft_p_t %>%
  group_by(Destination) %>%
  summarize(jobs = sum(Persons),
            .groups = "drop") %>%
  transmute(GTA06 = Destination,
            jobs)
```

Active travel modal splits:
```{r}
LTGO_split_ft <- od_ft_p_t %>%
  group_by(Origin) %>%
  summarize(LTGO_split = sum(T_LT, T_LTGO, T_GO)/sum(sum_trips)) %>% 
  transmute(GTA06 = Origin,
            LTGO_split)

W_split_ft <- od_ft_p_t %>%
  group_by(Origin) %>%
  summarize(W_split = sum(W)/sum(sum_trips)) %>%
  transmute(GTA06 = Origin,
            W_split)

BC_split_ft <- od_ft_p_t %>%
  group_by(Origin) %>%
  summarize(BC_split = sum(BC)/sum(sum_trips)) %>%
  transmute(GTA06 = Origin,
            BC_split)
```

Join workers and jobs to simple features object:
```{r}
ggh_taz <- ggh_taz %>%
  # Join workers
  left_join(workers_ft,
            by = "GTA06") %>%
  # Join jobs
  left_join(jobs_ft,
            by = "GTA06") %>%
  left_join(LTGO_split_ft,
            by = "GTA06") %>%
  left_join(W_split_ft,
            by = "GTA06") %>%
  left_join(BC_split_ft,
            by = "GTA06")
```

Replace NAs with 0s for the new columns:
```{r}
ggh_taz <- ggh_taz %>%
  mutate_at(vars(workers, jobs, LTGO_split, W_split, BC_split), ~replace_na(., 0))
```

## Visualize

@@

Plot job distribution in the ggh:
```{r}
ggplot() +
  geom_sf(data = ggh_taz,
          aes(fill = jobs),
          color = NA) +
  scale_fill_fermenter(direction = 1)
```
Plot transit split (0 to 1, 1 being 100% of trips made using transit) distribution in the ggh. Aggregating by CMA could be interesting:
```{r}
ggplot() +
  geom_sf(data = ggh_taz,
          aes(fill = LTGO_split),
          color = NA) +
  scale_fill_distiller(direction = 1)
```
Let's zoom into Toronto CMA, and plot the transit split
```{r}
#rotating the WGS84 crs so it displays Toronto aesthetically on a page; 17degree shift using +gamma
# lat_0 and lonc are the approximate centroid of Toronto
crs_string <- "+proj=omerc +lat_0=43.6532 +lonc=-79.3832 +alpha=0 +k_0=.7 +datum=WGS84 +units=m +no_defs +gamma=17"

#the bounding box, so we focus on Toronto
toronto_bb <- st_bbox(ggh_cma %>% filter(CMANAME %in% c("Toronto")))

tm_shape(ggh_taz, bbox = toronto_bb, projection=crs_string) +
  tm_polygons(col = "LTGO_split", border.alpha = 0.1, style = "jenks") +
  tm_shape(ggh_cma %>% filter(CMANAME %in% c("Toronto")), bbox = toronto_bb, projection=crs_string) +
  tm_borders()
```
The CMA/CAs overlapping all the TAZ:
```{r}
tm_shape(ggh_taz, projection=crs_string) +
  tm_polygons(col = "CMAUID", border.alpha = 0.1, style = "jenks") +
  tm_shape(ggh_cma, projection=crs_string) +
  tm_borders()
```

The planning boundaries overlapping all the TAZ:
```{r}
tm_shape(ggh_taz, projection=crs_string) +
  tm_polygons(col = "REGION", border.alpha = 0.1, style = "jenks") +
  tm_shape(ggh_pd, projection=crs_string) +
  tm_borders()
```

@@

## saving inputs for later processing
```{r}
save("ggh_taz", file = "TTS16-data-inputs/OD-by-FT-Employment/ggh_taz.Rdata")
save("ggh_pd", file = "TTS16-data-inputs/Boundaries/ggh_pd.Rdata")
save("ggh_cma", file = "TTS16-data-inputs/Boundaries/ggh_cma.Rdata")
save("od_ft_p_t", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_p_t.Rdata")
```

### 2022-10-22 data-raw/01-TTS16-Calculating-Travel-Time-OD.Rmd
additions to the TTS2016R package and minor edits to the manuscript (more work to be done).

Load:
```{r}
#Trips taken from origin to destination and associated people
load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft.Rdata")
```

```{r}
od_ft_tt <- od_ft %>% 
  merge(output_OD_car1, by.x=c("Origin", "Destination"), by.y=c("fromId", "toId"), all.x=T) %>%
  rename(trips = Persons)

rm(output_OD_car1)
summary(od_ft_tt)
```

@@

See '02-TTS-16-Imp-Function-Trip-Length-Distribution.Rmd' for further analysis:
```{r}
save("od_ft_tt", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_tt.Rdata")
```

>>>>

Load:
```{r}
# load(file = "TTS16-data-inputs/Travel-Time-Calculations/results/output_OD_car1.Rdata")
#Trips taken from origin to destination and associated people
load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_p_t.Rdata")
```

```{r}
od_ft_p_t_tt <- od_ft_p_t %>% 
  merge(output_OD_car1, by.x=c("Origin", "Destination"), by.y=c("fromId", "toId"), all.x=T)

rm(output_OD_car1)
summary(od_ft_p_t_tt)
```

@@

See '02-TTS-16-Imp-Function-Trip-Length-Distribution.Rmd' for further analysis:
```{r}
od <- od_ft_p_t_tt
rm(od_ft_p_t_tt)
save("od", file = "TTS16-data-inputs/OD-by-FT-Employment/od.Rdata")
```
